% ICML 2026 Submission
% Feedback Geometry: Topological Inconsistency Detection in Human Preference Data

\documentclass{article}

% For blind submission (comment out for camera-ready):
\usepackage{icml2026}
% For camera-ready, use:
% \usepackage[accepted]{icml2026}

% Standard packages
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{microtype}
\usepackage{xcolor}

% Theorem environments

ewtheorem{theorem}{Theorem}[section]

ewtheorem{proposition}[theorem]{Proposition}

ewtheorem{corollary}[theorem]{Corollary}

ewtheorem{lemma}[theorem]{Lemma}

ewtheorem{definition}[theorem]{Definition}

ewtheorem{remark}[theorem]{Remark}

% Running title
\icmltitlerunning{Feedback Geometry: Topological Inconsistency in RLHF}

\begin{document}

	wocolumn[
\icmltitle{Feedback Geometry: Topological Inconsistency Detection in Human Preference Data}

\icmlkeywords{Reinforcement Learning, Human Feedback, Sheaf Theory, Algebraic Topology, Hodge Theory, RLHF}

\begin{abstract}
Current Reinforcement Learning from Human Feedback (RLHF) strategies rely on an implicit ``scalar hypothesis'' that collapses complex, multidimensional human preferences into a single scalar value. This assumption fails when human preferences exhibit cyclic contradictions ($A \succ B \succ C \succ A$) or transitive inconsistencies. In this paper, we introduce a topological framework for analyzing preference data using Hodge decomposition on graph-based preference manifolds. By analyzing 160,800 Anthropic HH-RLHF prompt pairs, we identify that approximately 30\% of the data exhibits strong topological inconsistencies ($H^1 
eq 0$), making it mathematically impossible to represent within a standard scalar reward model without introducing significant bias or instability. We present a discrete Hodge-filtering approach that isolates the consistent gradient component from the inconsistent cyclic components. Our results demonstrate that Hodge-filtered reward models provide more stable training signals and prevent the generalization of cyclic biases that are otherwise exploitable by RL agents.
\end{abstract}

\vskip 0.3in
]

\printAffiliationsAndNotice{}

\input{sections/introduction}
\input{sections/background}
\input{sections/method}
\input{sections/experiments}
\input{sections/related_work}
\input{sections/conclusion}

\section*{Impact Statement}
This work provides tools for identifying and quantifying inconsistencies in human preference datasets used for AI alignment. By providing a rigorous mathematical framework for filtering non-transitive data, we enable the development of more stable and predictable RLHF systems.

\bibliography{references}
\bibliographystyle{icml2026}


ewpage
\appendix
\input{sections/appendix}

\end{document}
