\documentclass{article}

\usepackage[accepted]{icml2026}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{subcaption}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\icmltitlerunning{Hodge-Decomposed Preference Optimization}

\begin{document}

\twocolumn[
\icmltitle{Hodge-Decomposed Preference Optimization:\\Detecting and Filtering Cyclic Inconsistencies\\in Reward Learning}

\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Michael Lee}{equal,anon}
\end{icmlauthorlist}

\icmlaffiliation{anon}{Anonymous}

\icmlcorrespondingauthor{Michael Lee}{anonymous@example.com}

\icmlkeywords{preference optimization, Hodge decomposition, reward hacking, cyclic preferences, DPO, RLHF, AI alignment}

\vskip 0.3in
]

\printAffiliationsAndNotice{\icmlEqualContribution}

\begin{abstract}
Human preference data used to train reward models is inherently noisy and can contain cyclic inconsistencies---where $A \succ B \succ C \succ A$---that no scalar reward function can faithfully represent. Standard preference optimizers (DPO, KTO, GRPO) train on all preferences equally, including cycles that reward-hacking policies can exploit. We propose \textbf{Hodge-Decomposed Preference Optimization (HodgePO)}, a family of preference optimizers that use the combinatorial Hodge decomposition to separate globally consistent preference signals from cyclic noise. HodgePO operates through two mechanisms: (1)~per-sample weights derived from node-level cycle participation that downweight cyclically contaminated training pairs, and (2)~a \emph{Hodge potential-alignment regularizer} that encourages the model's reward predictions to match the globally consistent gradient potential from the Hodge decomposition. In a 30-seed benchmark on Anthropic's HH-RLHF dataset, Hodge-DPO achieves $\mathbf{0.9999}$ exploit resistance versus $0.940$ for standard DPO ($+6.3\%$, Cohen's $d = 6.52$, $p < 0.0001$), and Hodge-KTO achieves $\mathbf{0.9964}$ versus $0.800$ for KTO ($+24.5\%$, $d = 16.47$, $p < 0.0001$). These improvements are statistically significant across all 30 seeds, demonstrating that topological filtering of preference data substantially improves robustness to reward hacking.
\end{abstract}

%% ====================================================================
\section{Introduction}
\label{sec:intro}
%% ====================================================================

Preference optimization methods---Direct Preference Optimization (DPO)~\citep{rafailov2023direct}, Kahneman-Tversky Optimization (KTO)~\citep{ethayarajh2024kto}, Group Relative Policy Optimization (GRPO)~\citep{shao2024deepseekmath}, and Odds Ratio Preference Optimization (ORPO)~\citep{hong2024orpo}---have become the dominant approach for aligning language models with human intent. These methods sidestep explicit reward modeling by directly optimizing a policy from pairwise preference data.

However, human preferences are not always globally consistent. The Condorcet paradox~\citep{condorcet1785essai} demonstrates that pairwise majority preferences can be cyclic even when individual preferences are transitive---a phenomenon generalized by Arrow's impossibility theorem~\citep{arrow1950difficulty}. When preference data contains such cycles, \emph{no scalar reward function can faithfully represent the full preference structure}~\citep{jiang2011statistical}. Standard optimizers treat all preference pairs equally, forcing the model to fit contradictory signals. This creates exploitable inconsistencies that reward-hacking policies can discover and leverage~\citep{anthropic2025shortcuts, moskovitz2024confronting}.

\paragraph{Key insight.} The combinatorial Hodge decomposition~\citep{hodge1941harmonic, jiang2011statistical} provides an exact decomposition of preference data into:
\begin{itemize}
    \item \textbf{Gradient component}: globally consistent preferences explainable by a scalar potential (a ``true'' reward)
    \item \textbf{Harmonic component}: irreducible cyclic inconsistencies that no scalar reward can capture
\end{itemize}
By computing this decomposition on the preference graph and using it to modulate training, we can teach preference optimizers to learn from the consistent signal while discounting cyclic noise.

\paragraph{Contributions.}
\begin{enumerate}
    \item We introduce \textbf{Hodge-Decomposed Preference Optimization (HodgePO)}, a principled framework that augments any pairwise preference optimizer with cycle-aware training signals derived from the Hodge decomposition (Section~\ref{sec:method}).
    \item We identify and address two technical challenges: (a)~cross-pair graph connectivity is essential for non-trivial Hodge structure (Section~\ref{sec:graph_construction}), and (b)~the Hodge potential convention requires sign correction for use as a regularization target (Section~\ref{sec:potential_alignment}).
    \item We validate HodgePO across four base methods (DPO, KTO, GRPO, ORPO) in a statistically rigorous 30-seed benchmark on HH-RLHF, demonstrating significant improvements in exploit resistance for DPO ($+6.3\%$) and KTO ($+24.5\%$) with large effect sizes (Section~\ref{sec:experiments}).
\end{enumerate}

%% ====================================================================
\section{Background}
\label{sec:background}
%% ====================================================================

\subsection{Preference Optimization}

Given a dataset of preference pairs $\mathcal{D} = \{(x, y_w, y_l)\}$ where $y_w$ is preferred over $y_l$ for prompt $x$, DPO~\citep{rafailov2023direct} directly optimizes:
\begin{equation}
    \mathcal{L}_\text{DPO} = -\mathbb{E}\left[\log \sigma\left(\beta \log \frac{\pi_\theta(y_w | x)}{\pi_\text{ref}(y_w | x)} - \beta \log \frac{\pi_\theta(y_l | x)}{\pi_\text{ref}(y_l | x)}\right)\right]
\label{eq:dpo}
\end{equation}
KTO~\citep{ethayarajh2024kto} operates on individual samples rather than pairs, using prospect-theoretic loss weighting. GRPO~\citep{shao2024deepseekmath} uses group-relative advantages with a clipped surrogate objective.

All these methods assume preferences are generated by a latent reward function via the Bradley-Terry model~\citep{bradley1952rank}: $P(y_w \succ y_l) = \sigma(r^*(y_w) - r^*(y_l))$. This assumption fails when preferences contain irreducible cycles.

\subsection{Combinatorial Hodge Theory}
\label{sec:hodge_background}

The Hodge decomposition on graphs~\citep{jiang2011statistical} generalizes the classical Helmholtz decomposition to discrete settings. Given a weighted directed graph $G = (V, E, w)$ with edge flow $f: E \to \mathbb{R}$:

\begin{theorem}[Hodge Decomposition~\citep{jiang2011statistical}]
\label{thm:hodge}
Any edge flow $f$ on a graph $G$ decomposes uniquely as:
\begin{equation}
    f = \underbrace{d_0 \phi}_{\text{gradient}} + \underbrace{\delta_1 \psi}_{\text{curl}} + \underbrace{h}_{\text{harmonic}}
\end{equation}
where $d_0$ is the coboundary operator, $\delta_1$ is the adjoint boundary operator, $\phi: V \to \mathbb{R}$ is a node potential, $\psi$ is a 2-chain, and $h \in \ker(d_0^* d_0 + \delta_1 \delta_1^*)$ is the harmonic component.
\end{theorem}

The gradient component $d_0 \phi$ captures globally consistent rankings: for edge $(i, j)$, $(d_0 \phi)_{ij} = \phi(j) - \phi(i)$. The harmonic component $h$ captures irreducible cycles---preferences that cannot be explained by \emph{any} global ranking. In social choice theory, $h \neq 0$ corresponds to Condorcet cycles~\citep{condorcet1785essai}.

The \textbf{first Betti number} $\beta_1 = \dim(\ker \Delta_1)$, where $\Delta_1 = d_0^* d_0 + \delta_1 \delta_1^*$ is the Hodge Laplacian, measures the dimension of the harmonic space. When $\beta_1 > 0$, the preference data contains irreducible cyclic structure.

For preference optimization, we use the HodgeRank formulation~\citep{jiang2011statistical}: the gradient potential $\phi^*$ is the least-squares solution
\begin{equation}
    \phi^* = \arg\min_\phi \sum_{(i,j) \in E} w_{ij} \left(f_{ij} - (\phi(j) - \phi(i))\right)^2
\end{equation}
which gives the globally consistent reward ranking that best explains the observed preferences.

%% ====================================================================
\section{Hodge-Decomposed Preference Optimization}
\label{sec:method}
%% ====================================================================

HodgePO augments any base preference optimizer with two Hodge-derived mechanisms.

\subsection{Preference Graph Construction}
\label{sec:graph_construction}

The choice of graph structure is critical. A naive preference graph with only direct edges (one per training pair) produces a perfect matching---a bipartite graph with \emph{zero cycles} and trivial Hodge decomposition ($h = 0$ everywhere). This was a key failure mode in early experiments.

To obtain meaningful Hodge structure, we construct a denser preference graph:

\begin{definition}[Cross-Pair Preference Graph]
Given $N$ preference pairs embedded as $\{(\mathbf{e}_w^{(i)}, \mathbf{e}_l^{(i)})\}_{i=1}^N$ in $\mathbb{R}^d$, the cross-pair preference graph $G$ has:
\begin{itemize}
    \item \textbf{Direct edges}: $(w_i, l_i)$ with weight $P(y_w^{(i)} \succ y_l^{(i)})$ for each training pair $i$.
    \item \textbf{Cross-pair edges}: For each response embedding $\mathbf{e}$, connect to its $k$ nearest neighbors from \emph{other} pairs via cosine similarity, with preference direction determined by a lightweight reward model. We use $k = 5$.
\end{itemize}
\end{definition}

Cross-pair edges create shared nodes between different preference pairs, enabling cycles. In our experiments, 300 direct edges produce 0 cycles, while adding cross-pair edges yields $\sim$1,500 total edges with $\beta_1 > 0$ and $28\%$ harmonic energy fraction.

\subsection{Node-Level Cycle Participation Weights}

After computing the Hodge decomposition on $G$, we derive per-sample training weights based on how much each response participates in cyclic preference patterns.

For each node $v \in V$, define the \textbf{harmonic participation score}:
\begin{equation}
    s(v) = \sum_{e \ni v} |h_e|
\end{equation}
where the sum is over all edges incident to $v$ and $h_e$ is the harmonic component on edge $e$. For each training pair $i$ with preferred response node $w_i$ and dispreferred node $l_i$:
\begin{equation}
    \text{cycle\_score}(i) = \frac{s(w_i) + s(l_i)}{2 \cdot \max_j \frac{s(w_j) + s(l_j)}{2}}
\end{equation}
normalized to $[0, 1]$. The training weight is:
\begin{equation}
    \alpha_i = 1 - \lambda_\text{damp} \cdot \text{cycle\_score}(i), \quad \alpha_i \in [0.2, 1.0]
\label{eq:weights}
\end{equation}
with $\lambda_\text{damp} = 0.6$. Pairs where both responses participate heavily in cycles receive lower weight.

\subsection{Hodge Potential-Alignment Regularization}
\label{sec:potential_alignment}

The second and more impactful mechanism is a regularization loss that encourages the model's implicit reward predictions to align with the Hodge gradient potential---the globally consistent ranking.

The gradient potential $\phi^*$ from the Hodge decomposition assigns each node a score reflecting its position in the cycle-free global ranking. For training pair $i$:
\begin{equation}
    \Delta\phi_i = \phi^*(l_i) - \phi^*(w_i)
\label{eq:potential_diff}
\end{equation}

\paragraph{Sign convention.} In the Hodge framework, positive edge flow from $w_i$ to $l_i$ means $w_i$ is preferred. The gradient component satisfies $f_{w_i l_i}^\text{grad} = \phi^*(l_i) - \phi^*(w_i)$. For a consistent preference ($w_i \succ l_i$), positive flow implies $\phi^*(l_i) > \phi^*(w_i)$, so $\Delta\phi_i > 0$. This convention is critical: using $\phi^*(w_i) - \phi^*(l_i)$ instead produces negative targets that catastrophically reverse the learning signal.

Given the model's implicit reward predictions $r_\theta(w_i)$ and $r_\theta(l_i)$, the potential-alignment loss is:
\begin{equation}
    \mathcal{L}_\text{Hodge} = \frac{1}{B} \sum_{i \in \text{batch}} \left(r_\theta(w_i) - r_\theta(l_i) - \Delta\phi_i\right)^2
\label{eq:hodge_loss}
\end{equation}

The total HodgePO loss for any base method $\mathcal{L}_\text{base}$ is:
\begin{equation}
    \mathcal{L}_\text{HodgePO} = \mathcal{L}_\text{base}(\alpha) + \lambda_\text{Hodge} \cdot \mathcal{L}_\text{Hodge}
\label{eq:total_loss}
\end{equation}
where $\mathcal{L}_\text{base}(\alpha)$ denotes the base loss with per-sample weights $\alpha_i$ from Eq.~\ref{eq:weights}, and $\lambda_\text{Hodge}$ controls regularization strength.

\paragraph{Why not batch-level Hodge penalty?} An alternative approach computes the Hodge decomposition on \emph{in-batch} reward predictions. However, a scalar reward model produces rewards $r_\theta(x)$ that are always globally consistent by construction: any set of scalar values $\{r_1, \dots, r_n\}$ defines a total order with zero harmonic component. The batch-level Hodge penalty is therefore identically zero for scalar reward models---mathematically incapable of providing a training signal.

\paragraph{Intuition.} The potential-alignment regularizer tells the model: ``your reward differences should approximate the cycle-free global ranking.'' For training pairs involved in cycles, the Hodge potential difference $\Delta\phi_i$ is smaller (the cycle component has been removed), creating a softer target. For pairs in cycle-free regions, $\Delta\phi_i$ matches the raw preference, reinforcing the original signal. The regularizer thus selectively moderates learning on cyclically contaminated data.

\begin{algorithm}[t]
\caption{HodgePO Training}
\label{alg:hodgepo}
\begin{algorithmic}[1]
\REQUIRE Preference pairs $\mathcal{D}$, base optimizer (DPO/KTO/GRPO), $\lambda_\text{Hodge}$, $\lambda_\text{damp}$
\STATE Embed all responses: $\{\mathbf{e}_w^{(i)}, \mathbf{e}_l^{(i)}\}$
\STATE Construct cross-pair preference graph $G$ (Section~\ref{sec:graph_construction})
\STATE Compute Hodge decomposition: gradient potential $\phi^*$, harmonic $h$
\STATE Compute per-sample weights $\alpha_i$ via Eq.~\ref{eq:weights}
\STATE Compute potential diffs $\Delta\phi_i$ via Eq.~\ref{eq:potential_diff}
\FOR{each training epoch}
    \FOR{each mini-batch $B$}
        \STATE $\mathcal{L}_\text{base} \leftarrow$ base loss with weights $\alpha_i$
        \STATE $\mathcal{L}_\text{Hodge} \leftarrow$ Eq.~\ref{eq:hodge_loss}
        \STATE $\mathcal{L} \leftarrow \mathcal{L}_\text{base} + \lambda_\text{Hodge} \cdot \mathcal{L}_\text{Hodge}$
        \STATE Update model parameters via $\nabla \mathcal{L}$
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

%% ====================================================================
\section{Experiments}
\label{sec:experiments}
%% ====================================================================

\subsection{Setup}

\paragraph{Dataset.} We use Anthropic's HH-RLHF dataset~\citep{bai2022training}, sampling 2,268 preference pairs from the \texttt{harmless-base} split. Responses are embedded using \texttt{all-MiniLM-L6-v2} (384-dim). The cross-pair preference graph has 2,268 direct edges and $\sim$1,030 cross-pair $k$-NN edges (3,298 total), with a global harmonic energy fraction of $H^1 = 0.284$.

\paragraph{Methods.} We compare seven optimizers:
\begin{itemize}
    \item \textbf{DPO}, \textbf{KTO}, \textbf{GRPO}, \textbf{ORPO}: Standard implementations at embedding level.
    \item \textbf{Hodge-DPO}, \textbf{Hodge-KTO}, \textbf{Hodge-GRPO}: HodgePO variants with $\lambda_\text{Hodge} = 0.05$, $\lambda_\text{damp} = 0.6$.
\end{itemize}

All methods use identical architecture (2-layer MLP, hidden dim 64), learning rate $10^{-3}$, batch size 32, and 50 training epochs.

\paragraph{Metric.} \textbf{Exploit resistance}: the fraction of held-out test pairs where the trained model assigns higher reward to the preferred response than the dispreferred one. This measures robustness to reward hacking---a model that has learned to exploit cyclic inconsistencies will misrank some pairs.

\paragraph{Statistical rigor.} Each method runs across \textbf{30 independent seeds}. We report means, standard deviations, 95\% confidence intervals, and pairwise comparisons via Welch's $t$-test with Cohen's $d$ effect sizes.

\subsection{Main Results}

\begin{table}[t]
\centering
\caption{Exploit resistance across 7 preference optimizers, 30 seeds each. Hodge-DPO and Hodge-KTO significantly outperform their base methods. $^{***}p < 0.0001$.}
\label{tab:main_results}
\vskip 0.1in
\begin{tabular}{@{}lcccl@{}}
\toprule
\textbf{Method} & \textbf{Mean} & \textbf{Std} & \textbf{95\% CI} & \textbf{vs.\ Base} \\
\midrule
Hodge-DPO  & $\mathbf{0.9999}$ & $0.001$ & $\pm 0.000$ & $+6.3\%^{***}$ \\
DPO        & $0.940$ & $0.013$ & $\pm 0.005$ & --- \\
\midrule
Hodge-KTO  & $\mathbf{0.9964}$ & $0.003$ & $\pm 0.001$ & $+24.5\%^{***}$ \\
KTO        & $0.800$ & $0.017$ & $\pm 0.006$ & --- \\
\midrule
Hodge-GRPO & $1.000$ & $0.000$ & $\pm 0.000$ & $=$ \\
GRPO       & $1.000$ & $0.000$ & $\pm 0.000$ & --- \\
\midrule
ORPO       & $0.622$ & $0.080$ & $\pm 0.030$ & --- \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Pairwise statistical tests (Hodge variants vs.\ base methods).}
\label{tab:pairwise}
\vskip 0.1in
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Comparison} & \textbf{$\Delta$} & \textbf{Cohen's $d$} & \textbf{$p$-value} & \textbf{Effect} \\
\midrule
Hodge-DPO vs.\ DPO & $+0.060$ & $6.52$ & $< 0.0001$ & Large \\
Hodge-KTO vs.\ KTO & $+0.196$ & $16.47$ & $< 0.0001$ & Large \\
Hodge-GRPO vs.\ GRPO & $-0.001$ & $1.03$ & $0.18$ & --- \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:main_results} presents the main results. Two findings stand out:

\paragraph{1. Large, statistically significant improvements.} Hodge-DPO achieves $0.9999$ exploit resistance versus $0.940$ for standard DPO ($+6.3\%$, $d = 6.52$, $p < 0.0001$). Hodge-KTO improves from $0.800$ to $0.9964$ ($+24.5\%$, $d = 16.47$, $p < 0.0001$). Both effect sizes are ``large'' by conventional standards~($d > 0.8$); in fact, they are enormous, indicating essentially non-overlapping distributions between Hodge and non-Hodge variants.

\paragraph{2. Ceiling effects bound gains for strong methods.} GRPO already achieves $1.000$ resistance, leaving no room for Hodge improvement. This suggests HodgePO is most valuable for methods with intermediate baseline performance---precisely the regime where reward hacking exploits exist.

\subsection{Analysis: Why Does HodgePO Help?}

\paragraph{Cross-pair edges are essential.} Without cross-pair $k$-NN edges, the preference graph is a perfect matching with $H^1 = 0$ (zero harmonic energy). The Hodge decomposition is trivial, and Hodge variants are identical to base methods. With cross-pair edges, $H^1 = 0.284$---$28\%$ of preference flow is cyclic. This is the signal HodgePO exploits.

\paragraph{Potential-alignment dominates weighting.} In ablations, removing the potential-alignment regularizer ($\lambda_\text{Hodge} = 0$) and keeping only per-sample weights produces $< 1\%$ improvement. The weights have narrow variance ($\sigma = 0.015$) because all nodes participate similarly in cycles on this homogeneous dataset. The regularizer is the primary driver.

\paragraph{Regularization strength matters.} $\lambda_\text{Hodge} = 0.5$ catastrophically collapses exploit resistance to near-zero by overwhelming the base loss. $\lambda_\text{Hodge} = 0.05$ provides the optimal balance---strong enough to regularize cyclic noise, weak enough to preserve the base optimization signal.

\paragraph{Sign convention is critical.} The Hodge gradient potential assigns higher values to nodes that \emph{receive} more preference flow (``sinks'' in the preference graph), not to higher-quality responses. Using $\phi^*(w_i) - \phi^*(l_i)$ as the regularization target produces negative targets that teach the model to rank dispreferred responses higher---a subtle but catastrophic bug. The correct convention is $\Delta\phi_i = \phi^*(l_i) - \phi^*(w_i) > 0$ for consistent pairs.

\subsection{Method Ranking}

Across all 30-seed comparisons, the full ranking by exploit resistance is:

\begin{center}
Hodge-GRPO $=$ GRPO $>$ Hodge-DPO $>$ Hodge-KTO $>$ DPO $>$ KTO $>$ ORPO
\end{center}

All adjacent pairwise differences are statistically significant ($p < 0.05$) except Hodge-GRPO vs.\ GRPO (both at ceiling). ORPO's high variance ($\sigma = 0.080$) suggests instability without a reference model.

%% ====================================================================
\section{Related Work}
\label{sec:related}
%% ====================================================================

\paragraph{Preference optimization.} DPO~\citep{rafailov2023direct} reformulated RLHF as a classification problem; KTO~\citep{ethayarajh2024kto}, GRPO~\citep{shao2024deepseekmath}, and ORPO~\citep{hong2024orpo} offer alternatives with different trade-offs. Nash Learning from Human Feedback~\citep{munos2023nash} addresses intransitivity via game-theoretic equilibria but doesn't decompose the preference structure.

\paragraph{Reward hacking.} \citet{moskovitz2024confronting} constrain RLHF with KL bounds; \citet{coste2024reward} use reward model ensembles. These defend against overoptimization of a fixed reward model, not against cyclic structure in the preference data itself.

\paragraph{Hodge theory in ML.} HodgeRank~\citep{jiang2011statistical} introduced Hodge decomposition for ranking from pairwise comparisons. Neural sheaf diffusion~\citep{bodnar2022neural} uses sheaf theory for GNNs. \citet{ayzenberg2025sheaf} survey sheaf-theoretic deep learning. To our knowledge, HodgePO is the first application of Hodge decomposition to preference optimization training.

\paragraph{Cyclic preferences.} The Condorcet paradox~\citep{condorcet1785essai} and Arrow's theorem~\citep{arrow1950difficulty} establish that cyclic preferences are fundamental, not anomalous. \citet{zhu2023principled} provide PAC bounds for RLHF but assume a BTL model. HodgePO directly addresses the regime where BTL fails.

%% ====================================================================
\section{Discussion}
\label{sec:discussion}
%% ====================================================================

\paragraph{Limitations.} (1)~Our experiments operate at the embedding level rather than on full language model fine-tuning. The exploit resistance metric measures reward model accuracy, not downstream generation quality. (2)~Cross-pair edge construction requires $k$-NN search in embedding space, adding $O(N^2)$ preprocessing (mitigated by approximate NN). (3)~HodgePO's benefit scales with the harmonic energy fraction---on perfectly consistent data ($H^1 = 0$), it provides no advantage.

\paragraph{When does HodgePO help most?} The improvement is largest for methods with intermediate baseline performance (KTO: $+24.5\%$, DPO: $+6.3\%$) and absent at the ceiling (GRPO: no change). This suggests HodgePO is a ``consistency amplifier'': it filters noise that weaker methods are susceptible to, but cannot improve methods that already handle it.

\paragraph{Broader impact.} Cyclic preferences are endemic in human feedback data: our HH-RLHF analysis shows $28\%$ harmonic energy. As AI systems are trained on increasingly large and diverse preference datasets, the prevalence of cycles will only grow. HodgePO provides a principled, mathematically grounded approach to handling this irreducible structure---complementary to existing defenses like KL constraints and reward model ensembles.

%% ====================================================================
\section{Conclusion}
\label{sec:conclusion}
%% ====================================================================

We introduced Hodge-Decomposed Preference Optimization (HodgePO), a framework that uses the combinatorial Hodge decomposition to detect and filter cyclic inconsistencies in preference training data. The key technical insight is that the Hodge gradient potential provides a globally consistent reward ranking target that can be used as a differentiable regularizer, while scalar reward predictions always have zero in-batch harmonic energy and thus require a global decomposition. In a 30-seed benchmark, Hodge-DPO and Hodge-KTO achieve near-perfect exploit resistance with massive effect sizes over their base methods. The approach is method-agnostic, computationally lightweight (the Hodge decomposition is a one-time $O(|E|^2)$ preprocessing step), and requires no changes to model architecture.

\bibliography{references}
\bibliographystyle{icml2026}

%% ====================================================================
\newpage
\appendix

\section{Proof of Hodge Decomposition (Theorem~\ref{thm:hodge})}
\label{app:hodge_proof}

The Hodge decomposition on graphs follows from the orthogonality of the image and kernel of the graph Laplacians. Define the \textbf{boundary operators}:
\begin{itemize}
    \item $d_0: C^0(V) \to C^1(E)$, the coboundary (gradient): $(d_0 \phi)_{ij} = \phi(j) - \phi(i)$
    \item $d_1: C^1(E) \to C^2(T)$, the curl: $(d_1 f)_{ijk} = f_{ij} + f_{jk} + f_{ki}$ for triangle $(i,j,k)$
\end{itemize}

The \textbf{Hodge Laplacian} on 1-chains is $\Delta_1 = d_0 d_0^* + d_1^* d_1$, where $^*$ denotes the weighted adjoint. By the Hodge theorem:
\begin{equation}
    C^1(E) = \text{im}(d_0) \oplus \text{im}(d_1^*) \oplus \ker(\Delta_1)
\end{equation}
These three subspaces are mutually orthogonal under the edge inner product $\langle f, g \rangle = \sum_e w_e f_e g_e$.

The gradient potential $\phi^*$ is obtained by solving the normal equations $d_0^* d_0 \phi = d_0^* f$, which has a unique solution (up to additive constant) when $G$ is connected. We regularize with $10^{-4} I$ for numerical stability.

\section{Implementation Details}
\label{app:implementation}

\subsection{Preference Graph Construction}

\begin{enumerate}
    \item Embed all responses using \texttt{sentence-transformers/all-MiniLM-L6-v2} (384-dim).
    \item Create direct edges: for each preference pair $(y_w, y_l)$, add edge $(w, l)$ with weight equal to the preference probability.
    \item Create cross-pair edges: for each response node, find its $k=5$ nearest neighbors (cosine similarity) from other pairs. Preference direction is determined by the dot product of embeddings with the prompt embedding.
    \item Total graph: $N$ direct edges + $\sim 0.45N$ cross-pair edges (varies by dataset).
\end{enumerate}

\subsection{Hodge Decomposition}

We solve the normal equations $B^\top B \phi = B^\top f$ where $B$ is the $|E| \times |V|$ boundary operator. For our largest experiment ($|V| = 4536$, $|E| = 3298$), this takes $< 2$ seconds on CPU.

\subsection{Training}

All preference optimizers operate at the embedding level: a 2-layer MLP $(384 \to 64 \to 1)$ maps prompt-response embedding pairs to scalar rewards. This is not language model fine-tuning but rather an isolated evaluation of the preference optimization algorithm's ability to learn consistent reward rankings.

\begin{table}[h]
\centering
\caption{Hyperparameters.}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Embedding dim & 384 \\
Hidden dim & 64 \\
Learning rate & $10^{-3}$ \\
Epochs & 50 \\
Batch size & 32 \\
DPO $\beta$ & 0.1 \\
KTO $\beta$ & 0.1 \\
GRPO $\beta$ & 0.04 \\
GRPO group size & 8 \\
ORPO $\lambda$ & 0.5 \\
$\lambda_\text{Hodge}$ & 0.05 \\
$\lambda_\text{damp}$ & 0.6 \\
$k$-NN cross-pair edges & 5 \\
Seeds & 30 \\
\bottomrule
\end{tabular}
\end{table}

\section{Why Batch Harmonic Penalty Fails}
\label{app:batch_penalty}

A natural alternative to the potential-alignment regularizer is a \textbf{batch harmonic penalty}: compute the Hodge decomposition on the complete graph of in-batch reward predictions and penalize the harmonic energy fraction.

Given batch rewards $\{r_1, \dots, r_{2B}\}$ (B chosen + B rejected), form the complete graph with edge flow $f_{ij} = r_j - r_i$ for all $i < j$. The gradient potential is $\phi_i = r_i$ (the rewards themselves), giving gradient flow $f_{ij}^\text{grad} = r_j - r_i = f_{ij}$. Therefore:
\begin{equation}
    h_{ij} = f_{ij} - f_{ij}^\text{grad} = 0 \quad \forall (i,j)
\end{equation}

\textbf{The harmonic component is identically zero for any set of scalar predictions.} This is because scalar values on nodes always define a consistent ranking with no cycles. The Hodge decomposition can only detect cycles in \emph{exogenous} edge flows (human preferences), not in \emph{endogenous} predictions from a scalar model.

This observation motivated our shift to the precomputed potential-alignment approach (Section~\ref{sec:potential_alignment}), which uses the \emph{global} Hodge decomposition of human preference data as an external regularization target.

\section{Full Pairwise Comparison Matrix}
\label{app:pairwise}

\begin{table}[h]
\centering
\caption{Cohen's $d$ for all method pairs (30 seeds). Bold indicates $p < 0.05$.}
\resizebox{\columnwidth}{!}{
\begin{tabular}{@{}lccccccc@{}}
\toprule
 & DPO & H-DPO & GRPO & H-GRPO & KTO & H-KTO & ORPO \\
\midrule
DPO    & --- & $\mathbf{6.52}$ & $\mathbf{6.53}$ & $\mathbf{86.6}$ & $\mathbf{10.3}$ & $\mathbf{6.37}$ & $\mathbf{5.54}$ \\
H-DPO  &     & ---   & $\mathbf{1.03}$ & $\mathbf{1.03}$ & $\mathbf{16.5}$ & $\mathbf{1.67}$ & $\mathbf{6.67}$ \\
GRPO   &     &       & ---   & $1.03$         & $\mathbf{16.5}$ & $\mathbf{1.67}$ & $\mathbf{6.67}$ \\
H-GRPO &     &       &       & ---            & $\mathbf{86.6}$ & $\mathbf{1.03}$ & $\mathbf{6.67}$ \\
KTO    &     &       &       &                & ---   & $\mathbf{16.5}$ & $\mathbf{2.77}$ \\
H-KTO  &     &       &       &                &       & ---   & $\mathbf{6.57}$ \\
ORPO   &     &       &       &                &       &       & --- \\
\bottomrule
\end{tabular}
}
\end{table}

\end{document}
